Binary companions of red giant stars with APOGEE: \\ catalog and companions that survive the common envelope

### 2018-02-07 paper/1-catalog/twoface-catalog.tex
Cleaning up and various style edits

\newcommand{\nprior}{536,870,912}
\newcommand{\nposterior}{256}

@@

Time-domain radial-velocity measurements of stars contain information about
massive companions: even with two successive observations of a single star, a
difference in the measured radial velocities implies the existence of at least
one companion.
However, with few or imprecise radial-velocity measurements, the orbital
properties of the companion(s) are very poorly constrained (e.g.,
\citealt{Price-Whelan:2017}).
The vast majority of spectroscopic targets with repeat observations in the
largest (by number of objects) stellar spectroscopic surveys are often observed
just a few times with sparse, non-uniform phase coverage.
Most prior searches for companions using survey RV data have therefore
restricted their searches to only sources with many, high-precision epochs, so
that the orbital solution can be unambiguously determined (e.g.,
\citealt{Troup:2016}), or have used simple statistics computed from the data to
study multiplicity (e.g., ${\rm RV}_{\rm max} - {\rm RV}_{\rm min}$;
\citealt{Badenes:2017}).

@@

All data used in this work comes from the publicly-available data release 14
(\DR) of the \apogee\ survey (\citealt{apogeedr14}), a component of the Sloan
Digital Sky Survey IV (\sdssiv; \citealt{sdss,Blanton:2017}).
\apogee\ is designed to map stars across much of the Milky Way by obtaining
high-resolution ($R \sim 22500$) infrared ($H$-band) spectroscopy of primarily
\RGB\ stars.
Targets are selected with simple color and brightness cuts, but the survey uses
fiber-plugged plates with a maximum of 300 fibers per each $\approx 1.5~{\rm
deg}^2$ field of view, leading to ``pencil-beam''-like sampling of the stellar
distribution.
Data taken as part of the \apogee\ survey is reduced with the \apogee\ Stellar
Parameters and Chemical Abundances Pipeline (\acronym{ASPCAP};
\citealt{Garcia-Perez:2016}), which measures stellar parameters, chemical
abundances, and radial velocities for each target.
In order to meet signal-to-noise ratio requirements, most \apogee\ stars are
observed multiple times in a series of ``visits,'' typically with at least one
visit separated by a month or more in order to help identify binary stars.
\acronym{ASPCAP} iteratively solves for the relative velocities of each visit
spectrum while determining the optimal co-addition of the visit spectra.

We use the primary data products from \apogee\ \DR\ (i.e. the \texttt{allStar}
and \texttt{allVisit} files) which contain 258475 unique source IDs
(\texttt{APOGEE\_ID}) and 1054381 unique visits.
We select all stars with $\geq 3$ visits that each pass a set of quality cuts,
described below.
For each visit, we require that the visit velocity uncertainty is $< 100~\kms$
(\texttt{VRELERR}) and the following bits are not set in the \texttt{STARFLAGS}
bitmask: \texttt{PERSIST\_HIGH}, \texttt{PERSIST\_JUMP\_POS},
\texttt{PERSIST\_JUMP\_NEG}, \texttt{VERY\_BRIGHT\_NEIGHBOR}, \texttt{LOW\_SNR}.
For each star, we require that $0 < \log g < 3.5$ and the following bits are not
set in the \texttt{ASPCAPFLAGS} bitmask: \texttt{STAR\_BAD}.
After these cuts, and the requirement of $\geq 3$ visits for a given star, we
are left with \todo{XX} visits for \todo{XX} unique sources.
\figurename~\ref{fig:nvisits} shows the number of stars in several bins of
number of visits that pass the above quality cuts: \todo{XX}\% of the stars in
\DR\ have $< 8$ visits.

% Notebook: "Numbers after cuts"
\begin{figure}[hp]
\begin{center}
\includegraphics[width=\textwidth]{nvisits.pdf}
\end{center}
\caption{%
Number of \apogee\ stars in logarithmic bins of number of visits that pass the
quality cuts described in \sectionname~\ref{sec:data}.
In total, this work uses \todo{XX} visits and \todo{XX} unique sources.
\label{fig:nvisits}
}
\end{figure}

@@

For all \todo{XX} \apogee\ stars with $\geq 3$ good visits (see
\sectionname~\ref{sec:data}), we use \thejoker\ to generate posterior samplings
for each star under the assumptions listed above (see
\sectionname~\ref{sec:fitting}).
We start by generating a library of \nprior\ prior samples generated under the
prior defined in \citet{Price-Whelan:2017}: uniform or isotropic in angle
parameters, uniform in log-period over the domain $[1, 32768]~{\rm day}$, and a
beta distribution over eccentricity with fixed parameters
(\citealt{Kipping:2013}).
For the prior over the extra variance or jitter parameter, $s$, we initially use
a Gaussian over the transformed parameter $y = \ln s^2$ with the mean and
standard deviation of the initial distribution set to $(\mu_{y,0}, \sigma_{y,0})
= (10.6, 3)$ in units of ${\rm m}~{\rm s}^{-1}$.

We request \nposterior\ posterior samples for each source, and generate the
samples using an iterative procedure that adapts to the \todo{informativeness}
of the data to more efficiently generate posterior samples.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples.
When this happens, we proceed in one of two ways:
\begin{enumerate}
    \item \emph{needs MCMC}: If just 1 posterior sample is returned after
    exhausting the full library of prior samples, or if multiple (but fewer than
    \nposterior) are returned that all lie within a single mode of the posterior
    pdf, the orbit-space that is consistent with the data is treated as
    effectively unimodal. In this case, we use the location of the returned
    sample, or the median over returned samples to generate a small ball of
    initial conditions and use the ensemble MCMC sampler (\citealt{Goodman:})
    implemented in \python\ (\citealt{Foreman-Mackey:}) to continue sampling
    until we obtain \nposterior\ samples.
    \item \emph{needs more prior samples}: If more than 1 posterior sample is
    returned after exhausting the full library of prior samples and the samples
    lie in multiple modes of the posterior pdf, the star is flagged. In this
    case, the only way forward is to generate more prior samples and continue
    running \thejoker\ for these sources to generate more posterior samples.
\end{description}

A full run of \thejoker\ on all \todo{XX} \apogee\ stars used in this work took
approximately 300 hours on a compute cluster with 448 cores, with the time
dominated by sources with many ($\gtrsim 10$) visits.

@@

We use the per-source posterior samplings under the default prior described
above to infer the parameters of the assumed Gaussian distribution over the
log-jitter: $(\mu_y, \sigma_y)$.
This inference serves as a test-case for future work, where we intend to use the
independent posterior samplings to construct a hierarchical inference over
companion property distributions, and is also a test of the visit velocity
uncertainties reported in the \apogee\ data products.
If the catalog uncertainties are significantly underestimated, we expect the
inferred log-jitter distribution parameters to tend towards larger values.

To update the log-jitter distribution parameters, we maximize the marginal
likelihood of these population-level parameters.
We compute this marginal likelihood using the per-object posterior samples
re-weighted by the ratio of the value of the hyperprior evaluated at a new
location over the value of the default prior (see
\sectionname~\ref{sec:initsampling}).
This trick has been used in other hierarchical inferences as a way to
marginalize over the per-object parameters
(\citealt{Hogg:2008,Foreman-Mackey:...}), and we describe how to compute the
marginal likelihood in detail below.

\todo{Notation relative to above...}
For each $n$ of $N$ RGB stars in APOGEE, we obtain $K$ posterior samples over
primary orbital parameters $\bs{w} = (P, e, \omega, M_0, K, v_0)$ and the
jitter parameter, $y = \ln s^2$, using \thejoker.
To obtain this sampling, we use an interim (Gaussian) prior on the jitter
parameter parametrized by a mean and standard deviation $\alpha_0 = (\mu_{y,0},
\sigma_{y,0})$.

\begin{eqnarray}
    \bs{\theta}_{k} &=& (\bs{w}, y)_{k}\\
    \bs{\theta}_{k} &\sim& p(\bs{\theta}_k \given \{D_n\}, \alpha_0)
\end{eqnarray}

New hyperprior, parameters to infer: $\alpha = (\mu_y, \sigma_y)$.

We want to obtain posterior samples in
\begin{eqnarray}
    p(\alpha \given \{D_n\}) &\propto& p(\{D_n\} \given \alpha) \, p(\alpha)
\end{eqnarray}
where
\begin{eqnarray}
    p(\{D_n\} \given \alpha) &=& \prod_n^N p(D_n \given \alpha) \quad .
\end{eqnarray}

\begin{eqnarray}
    p(D_n \given \alpha) &=& \int \dd \theta_n \, p(D_n \given \theta_n) \,
      p(\theta_n \given \alpha)\\
    &=& \int \dd \theta_n \, p(D_n \given \theta_n) \, p(\theta_n \given \alpha) \,
      \frac{p(\theta_n \given D_n, \alpha_0)}{p(\theta_n \given D_n, \alpha_0)}\\
    &=& p(D_n \given \alpha_0) \, \int \dd \theta_n \,
      \frac{p(\theta_n \given \alpha)}{p(\theta_n \given \alpha_0)} \,
      p(\theta_n \given D_n, \alpha_0)\\
    &\approx& \frac{\mathcal{Z}_n}{K_n} \,
      \sum_k^{K_n} \frac{p(y_{nk} \given \alpha)}{p(y_{nk} \given \alpha_0)}
\end{eqnarray}

>>>>

\newcommand{\nprior}{536,870,912}
\newcommand{\nposterior}{256}
\newcommand{\nstars}{98,758}
\newcommand{\nvisits}{388,196}

@@

Time-domain radial-velocity measurements of stars contain information about
massive companions: even with two successive observations of a single star, a
difference in the measured radial velocities implies the existence of at least
one companion.
However, with few or imprecise radial-velocity measurements, the orbital
properties of the companion(s) are very poorly constrained (e.g.,
\citealt{Price-Whelan:2017}).
The vast majority of spectroscopic targets with repeat observations in the
largest (by number of objects) stellar spectroscopic surveys are often observed
just a few times with sparse, non-uniform phase coverage.
Most prior searches for companions using survey RV data have therefore
restricted their searches to only sources with many, high-precision epochs, so
that the orbital solution can be unambiguously determined (e.g.,
\citealt{Troup:2016}), or have used simple statistics computed from the data to
study multiplicity (e.g., $\textrm{RV}_\textrm{max} - \textrm{RV}_\textrm{min}$;
\citealt{Badenes:2017}).

@@

All data used in this work comes from the publicly-available data release 14
(\DR) of the \apogee\ survey (\citealt{apogeedr14}), a component of the Sloan
Digital Sky Survey IV (\sdssiv; \citealt{sdss,Blanton:2017}).
\apogee\ is designed to map stars across much of the Milky Way by obtaining
high-resolution ($R \sim 22500$) infrared ($H$-band) spectroscopy of primarily
\RGB\ stars.
Targets are selected with simple color and brightness cuts, but the survey uses
fiber-plugged plates with a maximum of 300 fibers per each $\approx
1.5~\textrm{deg}^2$ field of view, leading to ``pencil-beam''-like sampling of
the stellar distribution.
Data taken as part of the \apogee\ survey is reduced with the \apogee\ Stellar
Parameters and Chemical Abundances Pipeline (\acronym{ASPCAP};
\citealt{Garcia-Perez:2016}), which measures stellar parameters, chemical
abundances, and radial velocities for each target.
In order to meet signal-to-noise ratio requirements, most \apogee\ stars are
observed multiple times in a series of ``visits,'' typically with at least one
visit separated by a month or more in order to help identify binary stars.
\acronym{ASPCAP} iteratively solves for the relative velocities of each visit
spectrum while determining the optimal co-addition of the visit spectra.

We use the primary data products from \apogee\ \DR\ (i.e. the \texttt{allStar}
and \texttt{allVisit} files) which contain 258,475 unique source IDs
(\texttt{APOGEE\_ID}) and 1,054,381 unique visits.
We select all stars with $\geq 3$ visits that each pass a set of quality cuts,
described below.
For each visit, we require that the visit velocity uncertainty is $< 100~\kms$
(\texttt{VRELERR}) and the following bits are not set in the \texttt{STARFLAGS}
bitmask: \texttt{PERSIST\_HIGH}, \texttt{PERSIST\_JUMP\_POS},
\texttt{PERSIST\_JUMP\_NEG}, \texttt{VERY\_BRIGHT\_NEIGHBOR}, \texttt{LOW\_SNR}.
For each star, we require that $0 < \log g < 3.5$ and the following bits are not
set in the \texttt{ASPCAPFLAGS} bitmask: \texttt{STAR\_BAD}.
After these cuts, and the requirement of $\geq 3$ visits for a given star, we
are left with \nvisits\ visits for \nstars\ unique sources.
\figurename~\ref{fig:nvisits} shows the number of stars in several bins of
number of visits that pass the above quality cuts: \todo{XX}\% of the stars in
\DR\ have $< 8$ visits.

% Notebook: "Numbers after cuts"
\begin{figure}[hp]
\begin{center}
\includegraphics[width=\textwidth]{nvisits.pdf}
\end{center}
\caption{%
\todo{Update with new apogee db file}
Number of \apogee\ stars in logarithmic bins of number of visits that pass the
quality cuts described in \sectionname~\ref{sec:data}.
In total, this work uses \nvisits\ visits and \nstars\ unique sources.
\label{fig:nvisits}
}
\end{figure}

@@

For all \nstars\ \apogee\ stars with $\geq 3$ good visits (see
\sectionname~\ref{sec:data}), we use \thejoker\ to generate posterior samplings
for each star under the assumptions listed above (see
\sectionname~\ref{sec:fitting}).
We start by generating a library of \nprior\ prior samples generated under a
prior similar to that defined in \citet{Price-Whelan:2017}: uniform or isotropic
in angle parameters, uniform in log-period over the domain $[1,
32768]~\textrm{day}$, and a beta distribution over eccentricity with fixed
parameters (\citealt{Kipping:2013}).
For the prior over the extra variance or jitter parameter, $s$, we initially use
a Gaussian over the transformed parameter $y = \ln s^2$ with the mean and
standard deviation of the initial distribution set to $(\mu_{y,0}, \sigma_{y,0})
= (10.6, 3)$ in units of $\textrm{m}~\textrm{s}^{-1}$.

We request \nposterior\ posterior samples for each source, and generate the
samples using an iterative procedure that adapts to the \todo{informativeness}
of the data to more efficiently generate posterior samples.
For sources with very constraining data, \thejoker\ may return fewer than the
requested number of samples.
When this happens, we proceed in one of two ways:
\begin{enumerate}
    \item \emph{needs MCMC}: If just 1 posterior sample is returned after
    exhausting the full library of prior samples, or if multiple (but fewer than
    \nposterior) are returned that all lie within a single mode of the posterior
    pdf, the orbit-space that is consistent with the data is treated as
    effectively unimodal. In this case, we use the location of the returned
    sample, or the median over returned samples to generate a small ball of
    initial conditions and use standard MCMC to continue sampling until we obtain \nposterior\ samples. In detail, we use an ensemble sampler
    (\citealt{Goodman:}) implemented in \python\ (\citealt{Foreman-Mackey:}),
    run with \todo{XX} walkers for \todo{XX} steps, and downsample the resulting
    MCMC chains until we have \nposterior\ samples.
    \item \emph{needs more prior samples}: If more than 1 posterior sample is
    returned after exhausting the full library of prior samples and the samples
    lie in multiple modes of the posterior pdf, the star is flagged. In this
    case, the only way to proceed is to generate more prior samples and continue
    running \thejoker\ for these sources to generate more posterior samples.
\end{description}

A full run of \thejoker\ on all \nstars\ \apogee\ stars used in this work took
approximately 300 hours on a compute cluster with 448 cores, with the time
dominated by sources with many ($\gtrsim 10$) visits.

@@

As an initial use of the per-source posterior samplings, we use a hierarchical
Bayesian model to infer the parameters of the (assumed Gaussian) distribution
over the log-jitter parameter: $(\mu_y, \sigma_y)$.
This inference serves as a test-case for future work, where we intend to use the
independent posterior samplings to construct a hierarchical inference over
companion property distributions.
This is also a test of the visit velocity uncertainties reported in the \apogee\
data products: If the catalog uncertainties are significantly underestimated, we
expect the inferred log-jitter distribution parameters to tend towards larger
values.

In detail, we maximize the marginal likelihood of the population-level
parameters $\bs{\alpha} = (\mu_y, \sigma_y)$.
We compute this marginal likelihood using the per-object posterior samples
re-weighted by the ratio of the value of the hyperprior evaluated at a given
vector $\bs{\alpha}$ over the value of the default prior at the previously
assumed values, $\bs{\alpha}_0 = (\mu_{y,0}, \sigma_{y,0})$ (see
\sectionname~\ref{sec:initsampling}).
This trick has been used in other hierarchical inferences as a way to
marginalize over the per-object parameters
(\citealt{Hogg:2008,Foreman-Mackey:...}); we describe how to compute the
marginal likelihood in detail below.

For each $n$ of $N$ RGB stars in APOGEE, we obtain $K$ posterior samples over
primary orbital parameters $\bs{\theta} = (P, e, \omega, M_0, K, v_0)$ and the
jitter parameter, $y = \ln s^2$, using \thejoker; For brevity in expressions
below, we will use the vector
\begin{equation}
    \bs{w} = (\bs{\theta}, y)
\end{equation}
to represent the full set of parameters.
To obtain this sampling, we use an interim (Gaussian) prior on the jitter
parameter parametrized by a mean and standard deviation, i.e. $\alpha_0 =
(\mu_{y,0}, \sigma_{y,0})$ as described above.
For a given source, the posterior samples in the above parameters are drawn from
the distribution
\begin{equation}
    \bs{w}_k \sim p(\bs{w}_k \given D, \bs{\alpha}_0)
\end{equation}
where $D$ represents the data for a given object.

We want to compute the likelihood of all data from all $N$ stars, $\{D_n\}$,
given a new set of hyper-parameters $\bs{\alpha}$
\begin{equation}
    p(\{D_n\} \given \bs{\alpha}) = \prod_n^N p(D_n \given \bs{\alpha})
\end{equation}
where in the above, we have assumed that this likelihood is separable ( the data
for each source are independent).
The per-source marginal likelihood in the above expression is given by
\begin{align}
    p(D_n \given \bs{\alpha}) &= \int \dd \bs{w}_n \, p(D_n \given \bs{w}_n) \,
      p(\bs{w}_n \given \bs{\alpha})\\
    &= \int \dd \bs{w}_n \, p(D_n \given \bs{w}_n) \, p(\bs{w}_n \given \bs{\alpha}) \,
      \frac{p(\bs{w}_n \given D_n, \bs{\alpha}_0)}{p(\bs{w}_n \given D_n, \bs{\alpha}_0)}\\
    &= p(D_n \given \bs{\alpha}_0) \, \int \dd \bs{w}_n \,
      \frac{p(\bs{w}_n \given \bs{\alpha})}{p(\bs{w}_n \given \bs{\alpha}_0)} \,
      p(\bs{w}_n \given D_n, \bs{\alpha}_0) \label{eq:marglike} \quad .
\end{align}
Using the Monte Carlo integration approximation, \eqname~\ref{eq:marglike} can
be simplified to a sum over prior value ratios of the $K$ posterior samples for
each $n$ star
\begin{equation}
    \approx \frac{\mathcal{Z}_n}{K_n} \,
      \sum_k^{K_n} \frac{p(y_{nk} \given \bs{\alpha})}{p(y_{nk} \given \bs{\alpha}_0)}
      \quad .
\end{equation}
